On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-17 manuscripts/PNAS/chazthm_pnas.tex
discussion

We furthermore note that unlike the requirements of previous works \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}, the matrix $\mathbf{A}$ need not satisfy the spark condition to be recoverable from data. As a simple example for $k=2$, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ where $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$, and take $\mathcal{H}$ to be all consecutive pairs of $[m]$ arranged in cyclic order. Then, $\mathbf{A}$ satisfies the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} but not \eqref{SparkCondition}. %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\bar x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

@@

Before moving on to proofs, we discuss briefly the tightness of our number of required samples, $N$. We regard this as a basic open problem: how many samples determine the sparse linear coding model?  If $k$ is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in $\bar m$ and $k$, then a practical (polynomial) amount of data is enough to determine the model.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} shows that only a polynomial number of samples distributed over a polynomial number of supports suffices.} Reasons to be skeptical that a polynomial amount suffice in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}.

@@

We note that the explicit constants $C_1$ and $C_2$ are effectively computable. (The quantity $L_k$, may be computed as the smallest singular value of a certain matrix, while the quantity $\xi$ involves computing ``canonical angles'' between subspaces, which reduce again to an efficient singular value decomposition.) Because deciding $L_{k}(\mathbf{A}) > 0$ is already NP-hard \cite{tillmann2014computational}, it seems likely that without fixing a sparsity level there is no efficient computation $L_k$ in general; for fixed sparsity $k$, however, it is computable in polynomial time. Regardless, to calculate $C_2$ requires an exponential number of queries to $\xi$. Thus, $C_1$ and $C_2$ are not necessarily polynomial-time computable.

@@

\section{Discussion}\label{Discussion}
%[mention dictionary size independence]

% A significant implication is that the constant $C_1$ of Thm.~\ref{DeterministicUniquenessTheorem} can be computed efficiently if we restrict the class of possible supports for the $\mathbf{\bar x}_i$. 
Over the last twenty years, from the discovery of intrinsic ``Gabor'' wavelets in natural images \cite{donoho2001can} to the detection of forgeries \cite{hughes2010, Olshausen10}, 
sparse linear coding has become a standard tool in signal analysis, allowing for the unsupervised discovery of myriad insights into natural signals in a variety of scientific fields. 
%A compelling feature of this model is its simple instantiation of the principle of Occam's razor: a natural video, for instance, is modeled as a linear combination of a small number of spatiotemporal building blocks, each representing an archetypical feature latent in the data. 
For theoretical neuroscience in particular, dictionary learning for sparse linear coding and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006} that reproduce response properties of certain cortical neurons.  
In this work, we sought to understand when such universal representations be expected to emerge from the sparse linear coding model.
%to the recovery of ``mouse neuronal activity representing location on a track \cite{agarwal2014spatially}

%We have demonstrated the uniqueness and stability of solutions to Prob. \ref{InverseProblem} and the related Prob. \ref{OptimizationProblem}, 
To this end, we have conclusively verified that inference in this parsimonious model of (noisy) natural data is a well-posed problem in the sense of Hadamard \cite{Hadamard1902}, a long-standing necessary property of reductionist models in science. This notion summarizes the idea that inferences from observations should be robust to the inevitable uncertainty in measurement. In this regard, we demonstrate rigorously that, given sufficient data and suitable dictionary, Prob.~\ref{OptimizationProblem} reduces to an instance of Prob.~\ref{InverseProblem} to which applies our main result (Thm.~\ref{DeterministicUniquenessTheorem}): every dictionary and sparse codes solving Prob.~\ref{InverseProblem} are equivalent up to inherent relabeling/scaling ambiguities and a discrepancy (error) that scales linearly with the measurement noise or modeling inaccuracy; and we provide explicit, computable constants expressed in terms of these latent parameters.
%In other words, a small perturbation of the data should result in only slightly different inferred parameters. 
%Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise. 
%This work justifies the application of the sparse linear coding model to blind source separation problems, wherein the goal is to infer the generating dictionary and sparse codes from noisy measurements, as in \eqref{LinearModel}. We also collect a set of useful mathematical tools and basic facts for future research.

An immediate application of our theoretical work is Prop.~\ref{neuroprop}, which certifies the validity of perhaps the only published theory of neurally-plausible bottleneck communication between brain regions -- that unsupervised feature (dictionary) learning in a significantly smaller space of nervous activity can recover noisy sparse codes sent through a randomly-constructed (but unknown) wiring bottleneck \cite{Isely10}.  We refer the reader to \cite{ganguli2012compressed} for a discussion of this application's significance and more on the interactions between sparse linear coding and neuroscience.

One of the major innovations in our work is a theory of combinatorial designs for support sets (the ``singleton intersection property'') key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in combinatorial matrix theory (Lem.~\ref{MainLemma}) that will be useful in any noisy setting where uniqueness up to feature relabelling/scaling is desirable, and which draws upon the definition of a new matrix lower bound induced by a given hypergraph. Our arguments allow for the following theoretical advances: the dictionary size can be unknown; data require a polynomial, not exponential, number of distinct sparse supports and in certain settings a polynomial number of samples in total is sufficient; and the spark condition is not necessary for dictionary recovery.

A technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem} is the absence of any assumption on dictionaries in solutions to Prob.~\ref{InverseProblem}. Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure that an algorithm maintain a dictionary satisfying the spark condition \eqref{SparkCondition} at each iteration; indeed, even certifying a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.

@@

To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data. Moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}. In particular, a main reason for the sustained interest in dictionary learning for data analysis appears to be the assumed well-posedness of the model, confirmation of which forms the core of our  findings.

@@

%A reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} demonstrates that this result enables a practical amount of data to determine the model if one restricts the supports of the $\mathbf{\bar x}_i$ to a class that grows polynomially in $m, k$.

>>>>

We furthermore note that unlike in previous works \cite{li2004analysis, Georgiev05, Aharon06, Hillar15}, the matrix $\mathbf{A}$ need not satisfy the spark condition to be recoverable from data. As a simple example for $k=2$, let $\mathbf{A} = [ \mathbf{e}_1, \ldots, \mathbf{e}_5, \mathbf{v}]$ where $\mathbf{v} = \mathbf{e}_1 + \mathbf{e}_3 + \mathbf{e}_5$, and take $\mathcal{H}$ to be all consecutive pairs of $[m]$ arranged in cyclic order. Then, $\mathbf{A}$ satisfies the assumptions of Thm.~\ref{DeterministicUniquenessTheorem} but not \eqref{SparkCondition}. %(since $\{ \mathbf{A}_1, \mathbf{A}_3, \mathbf{A}_5, \mathbf{A}_6\}$ is not a linearly independent set).
%This weakening allows for a practical (polynomial) amount of data to still guarantee a stable dictionary in the case where the set of sparse supports for the $\mathbf{\bar x}_i$ is known to have a size that grows polynomially in $m$ and $k$.

@@

We close this section with a remark on sample complexity. A basic open problem remains: how many samples are necessary to determine the sparse linear coding model? If $k$ is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in $\bar m$ and $k$, then a practical (polynomial) amount of data suffices.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, tillmann2015computational}.

@@

The pragmatic reader should note that the explicit constants $C_1$ and $C_2$ are effectively computable. (The quantity $L_k$, may be computed as the smallest singular value of a certain matrix, while the quantity $\xi$ involves computing ``canonical angles'' between subspaces, which reduce again to an efficient singular value decomposition.) There is no efficient computation of $L_k$ in general, however, since even $L_{k} > 0$ is NP-hard \cite{tillmann2014computational}; although fixing $k$ yields polynomial complexity. Calculating $C_2$, however, requires an exponential number of queries to $\xi$; thus, $C_1$ and $C_2$ are not computable in polynomial time.

@@

\section{Discussion}\label{Discussion}

The goal of this work was to explain the emergence of consistent characteristic representations from sparse linear coding models fit to natural data despite the varied assumptions underlying the many practical methods in current use. To this end, we have taken an important step toward unifying the hundreds (if not thousands) of publications on the topic by demonstrating very general, deterministic conditions under which the identification of parameters in this parsimonious model is not only possible but also robust to the inevitable uncertainty permeating measurement and model choice.

Specifically, we have shown that, given sufficient data and suitable dictionary, the problem of seeking a dictionary and sparse codes with minimal average support size over the data (Prob.~\ref{OptimizationProblem}) reduces to an instance of Prob.~\ref{InverseProblem}, to which our main result (Thm.~\ref{DeterministicUniquenessTheorem}) applies: every dictionary and sparse codes solving Prob.~\ref{InverseProblem} are equivalent up to inherent relabeling/scaling ambiguities and a discrepancy (error) that scales linearly with the measurement noise or modeling inaccuracy. The constants we provide are explicit and computable; hence, there is an effective procedure sufficient to affirm if a proposed solution to Probs.~\ref{DeterministicUniquenessTheorem} or \ref{OptimizationProblem} is indeed unique (up to noise and inherent ambiguities).

An immediate application of our theoretical work is Prop.~\ref{neuroprop}, which certifies the validity of perhaps the only published theory to date of neurally-plausible bottleneck communication in the brain: that sparse coding via dictionary learning in a compressed space of nervous activity can recover sparse codes sent through a randomly-constructed (but unknown) noisy wiring bottleneck \cite{Isely10}.\footnote{We refer the reader to \cite{ganguli2012compressed} for a discussion of this application's significance and more on the interactions between sparse linear coding and neuroscience.}

%A compelling feature of this model is its simple instantiation of the principle of Occam's razor: a natural video, for instance, is modeled as a linear combination of a small number of spatiotemporal building blocks, each representing an archetypical feature latent in the data.

%For theoretical neuroscience in particular, dictionary learning and related methods have recovered characteristic components of natural images \cite{Olshausen96, hurri1996image, bell1997independent, van1998independent} and sounds \cite{bellsejnowski1996, smithlewicki2006} that reproduce response properties of certain cortical neurons.

%to the recovery of ``mouse neuronal activity representing location on a track \cite{agarwal2014spatially}

%Moreover, we show that even if the meta-parameter for the number of dictionary elements is overestimated, a subset of parameters may still be identifiable up to noise.

Beyond an original extension to the noisy regime, a major innovation in our work is a theory of combinatorial designs for support sets (the ``singleton intersection property'') key to the identification of the dictionary. We incorporate this idea into a new fundamental lemma in combinatorial matrix theory (Lem.~\ref{MainLemma}), which draws upon the definition of a new matrix lower bound induced by a hypergraph. Theoretical advances enabled by our combinatorial approach include: the dictionary size can be unknown and yet still a subset of parameters are recoverable; data require a polynomial, not exponential, number of distinct sparse supports and in certain settings a polynomial number of samples in total is sufficient; the dictionary need not satisfy the spark condition to be identifiable from data.

One technical difficulty in proving Thm.~\ref{DeterministicUniquenessTheorem} is the absence of any assumption on dictionaries in solutions to Prob.~\ref{InverseProblem}. Although mathematically interesting that no such requirement is necessary, there are other reasons to seek out such a theoretical guarantee. For instance, it is difficult to ensure that an algorithm maintain a dictionary satisfying the spark condition \eqref{SparkCondition} at each iteration; indeed, even certifying a dictionary has this property is likely intractable given its NP-hardness \cite{tillmann2014computational}.

@@

To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data. Moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}. In particular, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis appears to be the assumed well-posedness of parameter identification in the model, confirmation of which forms the core of our  findings.

%Since its inception some twenty years ago, sparse linear coding has become a standard tool in signal analysis, yielding myriad insights into the structure of natural signals across a variety of scientific domains. In this work,

@@
